---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The following code loads the data into R.  The data is stored in **Actdata**.

```{r loadData, results = "hide"}
        library(readr)
        Actdata <- read_csv("activity.zip")
```


## What is mean total number of steps taken per day?

### Total number of steps taken per day.

The data is grouped by day, then the total number of steps is calculated for each day and stored in the variable **sumSteps**.

```{r meanStepsDay}
        library(dplyr)
        
        ## Calculate the total number of steps by date
        sumStepsDay <-
                Actdata %>%
                group_by(date) %>%
                summarize(sum(steps))
        sumStepsDay

```

### Histogram of total number of steps taken each day

**sumStepsDay** is used to create a histogram of steps taken each day.

```{r histSteps}
        # Create a 'steps' column that is numeric
        sumStepsDay$steps <- as.numeric(sumStepsDay$`sum(steps)`)

        # Plot the histogram of steps taken each day
        hist(sumStepsDay$steps, col = "skyblue", xlab = "Total steps taken in a day", ylab = "Frequency (Days)", main = "Histogram of the number of steps taken each day")
        
```

The **mean** and **median** number of steps taken in a day can be calculated using the following which caters for the 'NA' values in the **steps** column.

Mean is `r mean(sumStepsDay$steps, na.rm = TRUE)`

```{r meanmedian}
        mean(sumStepsDay$steps, na.rm = TRUE)
        median(sumStepsDay$steps, na.rm = TRUE)
```


## What is the average daily activity pattern?
```{r meanStepsInterval}
        ## Calculate the average number of steps by interval
        avgStepsInt <-
                Actdata %>%
                group_by(interval) %>%
                summarize(mean(steps, na.rm = TRUE))
        avgStepsInt

        avgStepsInt$avgSteps <- as.numeric(avgStepsInt$`mean(steps, na.rm = TRUE)`)
        plot(avgStepsInt$interval, avgStepsInt$avgSteps, type = "l", col = "blue")
```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
